---
title: "Verbs `select` and `mutate`"
date: 'Last compiled: `r format(Sys.time(), "%b %d, %Y")`'
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r, label = "SETUP", echo = FALSE, results= 'hide', message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, comment = NA, fig.align = "center", fig.width = 4, fig.height = 4, message = FALSE, warning = FALSE)
library(dplyr)
library(hflights)
library(scales)
# Convert the hflights data.frame into a hflights tbl
hflights <- tibble::as_tibble(hflights)
# Create a random subset of size 10,000 to work with
set.seed(123)
train <- sample(227496, 10000)
hflights <- hflights[train, ]
# Create the object carriers, containing only the UniqueCarrier variable of hflights
carriers <- hflights$UniqueCarrier
# Both the dplyr and hflights packages are loaded into workspace
lut <- c("AA" = "American", "AS" = "Alaska", "B6" = "JetBlue", "CO" = "Continental", 
         "DL" = "Delta", "OO" = "SkyWest", "UA" = "United", "US" = "US_Airways", 
         "WN" = "Southwest", "EV" = "Atlantic_Southeast", "F9" = "Frontier", 
         "FL" = "AirTran", "MQ" = "American_Eagle", "XE" = "ExpressJet", 
         "YV" = "Mesa")

# Use lut to translate the UniqueCarrier column of hflights
hflights$UniqueCarrier <- lut[hflights$UniqueCarrier]
lut <- c("A" = "carrier", "B" = "weather", "C" = "FFA", 
         "D" = "security", "E" = "not cancelled")

# Use the lookup table to create a vector of code labels. 
# Assign the vector to the CancellationCode column of hflights
hflights$CancellationCode <- lut[hflights$CancellationCode]
```

_________________

## Verbs

The `dplyr` package contains five key data manipulation functions, also called verbs:

* `select()`, which returns a subset of the columns,
* `filter()`, that is able to return a subset of the rows,
* `arrange()`, that reorders the rows according to single or multiple variables,
* `mutate()`, used to add columns from existing data,
* `summarise()` or `summarize()`, which reduces each group to a single row by calculating aggregate measures.

If you want to find out more about these functions, consult the documentation by typing `?function_name` in the console. What order of operations should we use to to find the average value of the `ArrDelay` (arrival delay) variable for all American Airline flights in the  `hflights` `tbl`?

```{r}
OA <- filter(hflights, UniqueCarrier == "American" | UniqueCarrier == "American_Eagle")
head(OA) # Note only American flights ( Consider American and American Eagle to be the "American Airline")
mean(OA$ArrDelay, na.rm =  TRUE)
# Using the pipe operator to answer a slightly different question:
hflights %>% 
  group_by(UniqueCarrier) %>% 
  summarize(AD = mean(ArrDelay, na.rm = TRUE))
```

To answer the simple question whether flight delays tend to shrink or grow during a flight, we can safely discard a lot of the variables of each flight. To select only the ones that matter, we can use `select()`. Its syntax is plain and simple:

```{r, eval = FALSE}
select(data, Var1, Var2, ...) 
```

the first argument being the `tbl` you want to select variables from and the `VarX` arguments the variables you want to retain. You can also use the `:` and `-` operators inside of select, similar to indexing a `data.frame` with hard brackets. `select()` lets you apply them to names as well as integer indexes. The `-` operator allows you to select everything except a column or a range of columns.

## Instructions

* Use `select()` to return a copy of `hflights` that contains only the columns `ActualElapsedTime`, `AirTime`, `ArrDelay` and `DepDelay`.

```{r}
# Your code goes here
# Return a copy of hflights that contains the four columns related to delay.

select(ActualElapsedTime, AirTime, ArrDelay, DepDelay)

```

* Print `hflights` and assert that indeed, this dataset has not changed.

```{r}
# Your code goes here
# print hflights, nothing has changed!

```

* Return a copy of `hflights` containing the columns `Origin` up to `Cancelled`. Do not use integer indexes.

```{r}
# Your code goes here
# Return a copy of hflights containing the columns Origin up to Cancelled 
names(hflights)

```

* Find the most concise way to select: columns `Year` up to and including `DayOfWeek`, columns `ArrDelay` up to and including `Diverted`. You may want to examine the order of hflight's column names before you begin.

```{r}
# Your code goes here
# Answer to last question: be concise! 
names(hflights)


```

## Helper functions for variable selection

`dplyr` comes with a set of helper functions that can help you select variables. These functions find groups of variables to select, based on their names.  `dplyr` provides 6 helper functions, each of which only works when used inside `select()`.

* `starts_with("X")`: every name that starts with "X",
* `ends_with("X")`: every name that ends with "X",
* `contains("X")`: every name that contains "X",
* `matches("X")`: every name that matches "X", which can be a regular expression,
* `num_range("x", 1:5)`: the variables named `x01`, `x02`, `x03`, `x04` and `x05`,
* `one_of(x)`: every name that appears in `x`, which should be a character vector.

__Watch out:__ Surround character strings with quotes when you pass them to a helper function, but do not surround variable names with quotes if you are not passing them to a helper function.

## Instructions

* Use `select` and a helper function to return a `tbl` copy of `hflights` that contains just `ArrDelay` and `DepDelay`.

```{r}
#Your code goes here
# Return a tbl containing just ArrDelay and DepDelay


```

* Use a combination of helper functions and variable names to return the `UniqueCarrier`, `FlightNum`, `TailNum`, `Cancelled`, and `CancellationCode` columns of `hflights`.

```{r}
# Your code goes here
# Return a tbl as described in the second exercise, using both helper 
# functions and variable names


```

* Find the most concise way to return the following columns with `select` and its helper functions:  `DepTime`, `ArrTime`, `ActualElapsedTime`, `AirTime`, `ArrDelay`, `DepDelay`. Use only helper functions!

```{r}
# Your code goes here
# Return a tbl as described in the third exercise, using only helper functions.


```

## Comparison to basic `R`

To see the added value of the `dplyr` package, it is useful to compare its syntax with basic `R`. Up to now, you have only considered functionality that is also available without the use of `dplyr`. However, the elegance and ease-of-use of `dplyr` should be clear from this short set of exercises. To provide continuity, you will keep on working with the `hflights` dataset.

## Instructions

For exercises 1 to 3, duplicate each of the basic `R` commands shown by using dplyr-specific syntax. Don't use integer indexes!

```{r}
# Exercise 1
ex1r <- hflights[c("TaxiIn", "TaxiOut", "Distance")]
head(ex1r)
# Your code goes below


```

```{r}
# Exercise 2
ex2r <- hflights[c("Year", "Month", "DayOfWeek", "DepTime", "ArrTime")]
head(ex2r)
# Your code goes below


```

```{r}
# Exercise 3
ex3r <- hflights[c("TailNum", "TaxiIn", "TaxiOut")]
head(ex3r)
# Your code goes below


```

## Mutating is creating

`mutate()` is the second of five data manipulation functions you will get familiar with in this handout. In contrast to `select()`, which retains a subset of all variables, `mutate()` creates new columns which are added to a copy of the dataset.

Let's briefly recap the syntax:

```{r, eval = FALSE}
mutate(data, Mutant1 = expr(Var0, Var1,...))
```

Here, data is the `tbl` you want to use to create new columns. The second argument is an expression that assigns the result of any `R` function using already existing variables `Var0`, `Var1`, ... to a new variable `Mutant1`.

## Instructions

* Add a new variable named `ActualGroundTime` that measures the difference between `ActualElapsedTime` and `AirTime` to a copy of `hflights`. Save your results as `g1`.

```{r}
# Your code goes here
# Add the new variable ActualGroundTime to a copy of hflights and save the result as g1.


```

* Which variables in `g1` do you think count as a plane's "ground time"? Use `mutate()` to add these variables together and save them as `GroundTime`. Save your results as `g2`. Were you right? If you are, `GroundTime` should equal `ActualGroundTime` whenever an `NA` is not involved. Check this by printouts in the console.

```{r}
# Add the new variable GroundTime to a copy of g1 and save the result as g2.


```

* Add a new variable to `g2` named `AverageSpeed` that denotes the average speed that each plane flew in miles per hour. Save the resulting dataset as `g3`. Use `Distance / AirTime * 60`.

```{r}
# Your code goes here
# Add the new variable AverageSpeed to a copy of g2 and save the result as g3.


```

## Instructions

* A new variable `loss` is created with the first line of code below. Modify this code to create a second variable, `loss_percent`, the percentage of the `DepDelay` represented by `ArrDelay - DepDelay`. Assign the result to `m1`.

```{r}
m1 <- mutate(hflights, loss = ArrDelay - DepDelay)
# Your code goes here
# Add a second variable loss_percent to the dataset and save the result to m1.


```


* You will notice that the `m1` is redundantly defined: `loss_percent` uses `ArrDelay - DepDelay`, which is already defined as the `loss` variable. `mutate()` allows you to use a new variable while creating a second variable (or third variable, or etc.) in the same call. Rewrite the command above to use loss to define `loss_percent` and assign the result to `m2`.

```{r}
# Remove the redundancy from your previous exercise and reuse loss to define 
# the loss_percent variable.  Assign the result to m2.


```

* Using `mutate()`, create three new variables: (1) `TotalTaxi`, which is the sum of all taxiing times; (2) `ActualGroundTime`, which is the difference of `ActualElapsedTime` and `AirTime`; (3) `Diff`, the difference between the two newly created variables. Assign the result to `m3`. Observe that the `Diff` column will be zero at all times if you solved the exercise correctly!


```{r}
# Your code goes here
# Add the three variables as described in the third exercise and save the result to m3.


```

## Recap on `mutate` and `select`

As of now, you mastered two of the five data manipulation functions that are at the core of `dplyr`: `select()` and `mutate()`.

Which statement concerning the following four lines of code is correct?

A. `hflight <- select(hflights, -(Year:Month), -(DepTime:Diverted))`

B. `select(hflights, starts_with("D"))`

C. `select(hflights, -(Year:Month), -(DepTime:Diverted))`

D. `hflight <- select(hflights, starts_with("Day"))`


1. A. and C. lead to the same `hflight` variable and output; B. and D. do not.
2. A. and C. lead to the same `hflight` variable and output, as do B. and D.
3. A. and D. lead to the same `hflight` variable and output; B. and C. do not.
4. A. and D. lead to the same hflights variable and output, as do B. and C.


```{r}
# Your verification code goes here


```

______________

